From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-03-06 paper/paper.Rmd
better creator section intro, new conclusion section

We choose this model because it has precisely same functional form
as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
but is explicit about how this net growth rate is divided between
birth and death rate.  As we shall see, this is not merely a notational
convenience but leads to real differences in observed stochastic properties
of the model. Both the numerical approach (details in Appendix A) and the 
analytical approach (Appendix B) illustrate that it is straight-forward
to use this same approach in alternate formulations.

Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$, illustrating the magnitude of these fluctuations indeed scales in proportion
to $\sqrt{N}$ as postulated by the canonical equation, Eq \eqref{canonical}.

@@

The van Kampen system size expansion provides a mechanism to approximate the individual
dynamics represented by the master equation (equivalently, by the Gillespie simulation)
with a population-level description of the process.  This is important not only because
it will illustrate both origins and approximations that lie behind the population level
models that predominate this literature, but will give our syntheses a chance to point
out some more subtle misconceptions that have also arisen. Details of this expansion
for a generic birth death process and comparison with the more familiar diffusion 
approximation are found in Appendix B.

van Kampen's approximation follows a central-limit theorem style argument established rigorously in
the theorems of Kurtz at about the same time [e.g @Kurtz1970; @Kurtz1972; @Kurtz1978; theorem 4.4.].
The heart of the approximation is to imagine a population on some large landscape of which
we sample only a fixed area to estimate the population density.  That area is our 
system size, which as it expands includes more and more total individuals and
experiences proportionally smaller fluctuations in population density.  This 
results in the conclusion that the population density obeys a simple Gaussian
distribution, whose mean size $x$ and variance  $\sigma^2$ are given by *ordinary* differential equations:

@@

This result is striking for a variety of reasons.  First, we see the recovery of the 
deterministic skeleton, in which we naively replace population density $x/N$ for
population count $n$ describes the *expected* density of the population
(i.e. averaged over an ensemble of replicate realizations of the Gillespie simulations),
and with remarkable accuracy, up to order $1/N$, i.e. a single individual.^[We can actually extend the van Kampen expansion to higher order and successfully approximate even this higher-order correction to the mean dynamics, recovering the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model depending on the curvature of the macroscopic equation].
The variance term is equally interesting.  First, we note that noise arises from
the interaction of individual birth and death events that defined our process,
illustrating the origins of this noise.  Because both individual events of births
and deaths add randomness, they both add to the growth of variation observed, i.e.
$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma^2$ are *non-linear* whenever
$b(x)$ or $d(x)$ are non-linear, so that different values of the population density
will experience different noise intensities! In fact, this variance can only
have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] < 0$)
Third, this expansion shows the variance
equation is also less precise, accurate only to order $1/\sqrt{N}$. 
This variance equation will be important to us later as we examine the relative
roles of noise and non-linearity in noise-driven phenomena such as stochastic
transitions between alternate stable states.

The expected auto-correlation, $\langle x_0, x_t \rangle$ also follows from the
system size expansion, e.g. at stationary state:

@@

This better-known approach has several potential draw-backs.  First, as noted in @vanKampen2007 or
@Gardiner2009 (see Appendix B), the Kramer-Moyal expansion approach rests on rather
a more dubious double limit that often has no clear biological interpretation and
can restrict its validity (though I believe theorem 3.3 of @Kurtz1978 establishes
this approximation on firmer grounds).  Second, the non-linear SDE formulation can
prove more difficult to work with than the convenient ODE expressions from van Kampen,
Eq \eqref{vanKampen} (which can be expressed as a coupled ODE for the mean and 
a linear SDE for deviations).  Third, while both approximations break down
in the case of large deviations [see @Ovaskainen2010], the van Kampen expansion
shows that the macroscopic equation will continue to hold even when the 
variance does not; in contrast to the non-linear SDE from the diffusion approximation.

@@

These examples of how a stochastic model arises from both intrinsic and extrinsic
sources of noise through the use of the system size expansion serve both
as critique and justification of the familiar notion of a stochastic model as 
a deterministic skeleton plus some Gaussian white noise term.  Critique, because
the expansions show that both noise and deterministic skeleton are the 
result of lower-level mechanistic processes, whose precise form will 
depend on the details of those processes.  Yet the expansion also shows us that
this Gaussian noise description can be a surprisingly good approximation,
that noise intensity will converge to a stationary value whenever the system
dynamics have a stable point, and that the contribution of extrinsic
(i.e. environmental) noise factors will dominate in large systems.  These
justifications will allow us to pursue our analysis of noise-driven phenomena
using simple models in which a nonlinear deterministic skeleton is driven by
Gaussian environmental noise.

@@

Outside of the primarily theoretical literature on stochastic dynamics
in ecology, noise is viewed largely as a nuisance: something entirely
distinct from the processes we are interested in which invariably keeps
getting in the way.  As we have seen in section on origins, such 
characterizations are misleading: noise arises from the very same processes.
While both empirical and statistical literature have frequently emphasized
techniques to remove noise, a central theme of stochastic processes
have emphasized an almost opposite conclusion: noise cannot be subtracted or
averaged away, since the presence of noise can fundamentally alter the
dynamics we observe.
Stochasticity can give rise to a wide array of interesting phenomena
that are often at odds with our intuition. As these examples will
illustrate, such behavior arises through an interaction between
stochastic and non-linear elements of a system, underscoring the
importance of explicitly including stochasticity in ecological
models [@Coulson2004; @Higgins1997].  Importantly, the noise itself
does not have to be large to give rise to these phenomena.  In this
section, I highlight several examples that have shown up frequently
in the literature and prove sufficiently tractable to provide a
simple intuition for the mechanisms involved.

@@

The loss of a species is without a doubt one of the most important
possible consequences of noise. Unsurprisingly few topics have
received greater attention then the question of the persistence of
a given species [e.g. @Kendall1949;  @Bartlett1955; @Bartlett1960;
@Lande1993; @Saether1998; @Coulson2001; @Schreiber2010] and coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017]
in stochastic models.  Two important trends in this literature have
been establishing existing results in more general and precise language
and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
overview and synthesis of persistence and coexistence respectively],
and continuing to extend such results to more complex contexts 
[@Ovaskainen2010 provides one excellent review], in
particular addressing auto-correlated noise [e.g. @Ripa; @Ripa1999; 
@Petchey; @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
[e.g. @Durrett1994; @Hastings1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
among other issues.

Demographic noise in particular has fundamental implications for the long-term
equilibrium: any finite population, however big, is guaranteed to
eventually go extinct under demographic noise [e.g. @Kendall1949; @Lande2003].  
In most models it is possible [@Schreiber2006] to define a notion of quasi-stationary 
distribution concentrated around the deterministic attractor,
and to prove the probability of extinction decreases exponentially with 
system size [@Ovaskainen2010; @Schreiber2017]. We will turn our attention
away from the well-trodden ground of persistence and coexistence to focus
on other phenomena that can emerge in stochastic models.  This will
highlight the importance of noisy phenomena beyond the ability to obscure
a deterministic skeleton of some underlying process or in driving extinctions,
but rather in shaping and driving other interesting patterns we observe in
natural systems. It is these phenomena that will also underpin the third
transition in which noise can be turned to our advantage as a source
of information about ecological processes.

@@

Despite the nearly ubiquitous association of a ball-in-cup model
with ecological discussions of resilience and stability, this
potential well diagram (Figure \ref{noisy_switch}C) is also a source of confusion
on the topic.  As we have seen, the most salient feature of this
diagram is the depth of each well: transition probabilities (under
small noise) depend only on the depth of the well and not the
steepness of the well, the distance between the wells or between
well and tipping point barrier.  However, it is important to remember
that because the potential function is defined by the above integral,
it in fact already combines two more elementary and mechanistically
precise quantities from the growth equation: the slope of the growth
rate (stability) and the distance to the tipping point (size of the
basin of attraction).  Both a steeper negative slope or a longer
distance to integrate over between stable and unstable points will
create a deeper well.

Because these transitions between alternative stable states appear as
such a dramatic deviation from the previous pattern (as illustrated
in Figure \ref{noisy_switch}A), it is  tempting to assume they
result from some sudden external force or change in the
environment. As the numerical examples illustrate,
these transitions can be explained entirely by chance
[@Boettiger2012;  @Drake2013].

@@

The trick of the early warning signals approach is to view noise as 
a sequence of countless miniature perturbation experiments.  
Figure \ref{tipping} illustrates early warning indicators in an
experimental manipulation and corresponding simulated model of
@Dai2012 and @Dai2015.  Though early warning signals have been
reported in a growing range of systems [@Scheffer2015], these
laboratory experiments in population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cerevisiae_ on a sucrose medium provide the most convincing 
mechanistic interpretation of the underlying fold bifurcation.
Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation which is accompanied
by the patterns of increasing variance and increasing auto-correlation.
This pattern can be seen empirically in lab populations [@Dai2012; @Dai2015]
and explored numerically in a simulation of the yeast dynamics shown in 
Figure \ref{tipping}. As the mechanistic model of yeast population
dynamics is already quite simple, the simulation shown here (code in Appendix B)
corresponds directly to the model proposed by @Dai2015.

@@

The potential for stochastic phenomena to provide information about underlying
processes goes well beyond avialable examples in stability and early warning signals.
The examples from fold bifurcations already suggest
that stochastic processes could play a similar role in identifying
other critical transitions. For instance, stochastic resonance may be 
visible in advance of a Hopf bifurcation, where a stable state gives
way to stable oscillations (limit cycle) of growing magnitude
[e.g. @Fussmann2000; @Nelson2013; @Strogatz2001].  Noise can provide
information in unrelated ways as well. in deriving expressions for the variance created by
demographic noise above, we observed that different partitions of
a logistic growth model can lead to the same macroscopic dynamics
(same values for $b$ - $d$), while having different variance (due
to different values for $b$+$d$).  This suggests that the variation
observed in such a population could potentially help resolve
how dynamics are partitioned between individual birth and death
rates (though in practice one would have to first account for any
contribution of environmental noise).

@@

The past two decades have seen widespread recognition of the central
role that stochasticity plays through the use of mechanistic models
can have in creating and understanding ecological phenomena
[@Coulson2004; @Black2012]. We have seen increasing use of
master-equation approaches to individual-based models to reflect
demographic stochasticity that is amenable to analytical treatment
[@Black2012], and much work in understanding the importance of
environmental noise, particularly the more realistic but less
mathematically convenient notion of colored or auto-correlated noise
on population dynamics, persistence, and coexistence [@Ovaskainen2010;
@Schreiber2017].  These deductive models can generate phenomena
that resemble familiar patterns such as population cycles for which
we have previously sought deterministic explanations, as well as
stochastic phenomena with no such analog such as stochastic switching.
Yet too many of these advances have remained limited to the domain
of theory; limited to deductive models which may seek to mimic
observed patterns but are rarely used to confront empirical data
directly.  There is good reason for this: statistical inference of
even relatively simple models with multiple sources of noise will
always be difficult given limited ecological data [@Knape2011].
However, many stochastic phenomena may suggest new ways to infer
rather than merely obscure underlying processes.

The eruption of interest in early warning signals for critical
transitions [@Scheffer2009] provides an excellent illustration of
this potential.  A range of stochastic phenomena have been suggested
and tested for their potential to reveal subtle signs of ecosystem
change prior to a dramatic shift, and prompted a rich dialog between
further theory and new experiment. Changes in stochastic properties
such as variance or auto-correlation provide evidence of resilience
loss that would be invisible to a deterministic model, and non-linear
stochastic phenomena such as stochastic switching are observed as
evidence of 'flickering' prior to a critical transition.  In this
way, our understanding of stochastic dynamics is translated into
novel sources of information about a system which has sparked a
wealth of empirical investigation and further theoretical work to
test the potential of this approach.  The study and application of
stochastic phenomena to detect ecosystem change is unlikely to be
limited to this simple scenario of a critical transition or fold
bifurcation.

Precisely because most stochastic phenomena arise through their
interaction with non-linearity, the patterns that result can become
a lens into better understanding those nonlinear processes and
detecting how they may change.  Meanwhile, the widespread adoption
of computational tools and the ability to express theoretical results
in numerical algorithms can make any new approaches more readily
accessible to application in empirical research. In the past two
decades, we have set to rest those deterministic skeletons that see
noise only as nuisance and embraced the central role noise plays
as the creator of ecological phenomena. Going forward, I hope and
predict that we will see greater discussion and use of stochastic
phenomena to infer underlying processes in empirical patterns in
similar interactions between theoretical and empirical research.

>>>>

We choose this model because it has precisely same functional form
as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
but is explicit about how this net growth rate is divided between
birth and death rate.  As we shall see, this is not merely a notational
convenience but leads to real differences in observed stochastic properties
of the model. Both the numerical approach (details in Appendix A) and the 
analytical approach (Appendix B) illustrate that it is straight-forward
to use this same approach in alternate formulations.
Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$, illustrating the magnitude of these fluctuations indeed scales in proportion
to $\sqrt{N}$ as postulated by the canonical equation, Eq \eqref{canonical}.

@@

The van Kampen system size expansion provides a mechanism to approximate the individual
dynamics represented by the master equation (equivalently, by the Gillespie simulation)
with a population-level description of the process.  This is important not only because
it will illustrate both origins and approximations that lie behind the population level
models that predominate this literature, but will give our syntheses a chance to point
out some more subtle misconceptions that have also arisen. Details of this expansion
for a generic birth death process and comparison with the more familiar diffusion 
approximation are found in Appendix B.  van Kampen's approximation follows a central-limit
theorem style argument established rigorously in
the theorems of Kurtz at about the same time [@Kurtz1970; @Kurtz1972; @Kurtz1978].
The heart of the approximation is to imagine a population on some large landscape of which
we sample only a fixed area to estimate the population density.  That area is our 
system size, which as it expands includes more and more total individuals and
experiences proportionally smaller fluctuations in population density.  This 
results in the conclusion that the population density obeys a simple Gaussian
distribution, whose mean size $x$ and variance  $\sigma^2$ are given by *ordinary* differential equations:

@@

This result is striking for a variety of reasons.  First, we see the recovery of the
deterministic skeleton, in which we naively replace population density $x/N$ for
population count $n$ describes the *expected* density of the population
(i.e. averaged over an ensemble of replicate realizations of the Gillespie simulations),
and with remarkable accuracy, up to order $1/N$, i.e. a single individual.^[We can actually extend the van Kampen expansion to higher order and successfully approximate even this higher-order correction to the mean dynamics, recovering the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model depending on the curvature of the macroscopic equation].
The variance term is equally interesting.  First, we note that noise arises from
the interaction of individual birth and death events that defined our process,
illustrating the origins of this noise.  Because both individual events of births
and deaths add randomness, they both add to the growth of variation observed, i.e.
$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma^2$ are *non-linear* whenever
$b(x)$ or $d(x)$ are non-linear, so that different values of the population density
will experience different noise intensities! In fact, this variance can only
have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] < 0$)
Third, this expansion shows the variance
equation is also less precise, accurate only to order $1/\sqrt{N}$. 
This variance equation will be important to us later as we examine the relative
roles of noise and non-linearity in noise-driven phenomena such as stochastic
transitions between alternate stable states. The expected auto-correlation,
$\langle x_0, x_t \rangle$ also follows from the
system size expansion, e.g. at stationary state:

@@

This better-known approach has several potential draw-backs.  First, as noted in @vanKampen2007 or
@Gardiner2009 (see Appendix B), the Kramer-Moyal expansion approach rests on rather
a more dubious double limit that often has no clear biological interpretation and
can restrict its validity (though I believe theorem 3.3 of @Kurtz1978 establishes
this approximation on firmer grounds).  Second, the non-linear SDE formulation can
prove more difficult to work with than the convenient ODE expressions from van Kampen,
Eq \eqref{vanKampen} (which can be expressed as a coupled ODE for the mean and 
a linear SDE for deviations).  Third, while both approximations break down
in the case of large deviations [see @Ovaskainen2010], the van Kampen expansion
shows that the macroscopic equation will continue to hold even when the 
variance does not; in contrast to the non-linear SDE from the diffusion approximation.
Though the limit used in the diffusion approximation can break down in cases where the 
van Kampen expansion is still valid [see @Gardiner2009], in most circumstances they
produce similar results and both justify the postulates of the Eq \eqref{canonical}.

@@

These examples of how a stochastic model arises from both intrinsic and extrinsic
sources of noise through the use of the system size expansion serve both
as critique and justification of the familiar notion of a stochastic model.
This approximation justifies pulling out the stochasticity and adding it on the end
of a deterministic skeleton (van Kampen's macroscopic equation) as Gaussian 
white noise term.  The expansion also serves as a reminder that
both noise and deterministic skeleton are the 
result of lower-level mechanistic processes, whose precise form will 
depend on the details of those processes. The Gaussian form, though
convenient, is still an approximation, not suitable for small populations
or rare, large deviations [see Appendix B and @Ovaskainen2010].
The expansion has given us three things: (1) showing us that additive
Gaussian noise can be a surprisingly good approximation, (2)
that noise intensity will converge to a stationary value whenever the system
dynamics have a stable point, and (3) that the contribution of extrinsic
(i.e. environmental) noise factors will dominate in large systems.  These
justifications will allow us to pursue our analysis of noise-driven phenomena
using simple models in which a nonlinear deterministic skeleton is driven by
Gaussian environmental noise.

@@

Most ecologists first encounter stochastic models in the context of statistics,
where noise is just a nuisance to be stripped away [@Gotelli2004; @Bolker2007]. 
In examining its origins, we were able to justify the simplistic
view of noise as something merely added on to a deterministic skeleton
to be less ad hoc than it appears; justified as it is by careful
approximations and underlying theorems so long as our systems are 
sufficiently large and the fluctuations sufficiently small.  If this
appears to suggest that noise is somewhat negligible after all as long
as we have the deterministic skeleton, then this section is all about 
rejecting that notion.

A major theme of the literature on stochastic models has been the illustration
that, even in large systems with small noise, the qualitative behavior
of a stochastic model can be entirely unlike the behavior of the deterministic
skeleton [@Higgins1997; @Bjornstad2001; @Coulson2004].  Even with our
focus on population dynamics alone, examples of these differences are too
numerous to do justice to all of them here -- instead, we will focus on
a handful of examples that show how noise interacts with non-linearity
to create phenomena that are absent from the deterministic model. These examples
will not only illustrate how ecological theory often sees noise in a
very different role from ecological statistics, but will serve us well
in our third section which seeks to turn that view of noise entirely on 
it's head: not as something that obscures underlying processes but rather as
a way of revealing them.

@@

Perhaps there is no starker example of differences between the behavior
of a deterministic skeleton and a stochastic model than in the case
of persistence of a species and the coexistence of multiple species.  
Demographic noise guarantees extinction in the long-run for any finite
population [e.g. @Kendall1949; @Lande2003;  @Bartlett1955; @Bartlett1960;
@Lande1993; @Saether1998; @Coulson2001; @Schreiber2006;  @Schreiber2010; @Ovaskainen2010;
@Schreiber2017]. Stochasticity can also
do the opposite, allowing many species to coexist in scenarios where
the deterministic skeleton would predict all but one of them to be
doomed to extinction [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017].
Two important trends in this literature have
been establishing existing results in more general and precise language
and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
overview and synthesis of persistence and coexistence respectively],
and continuing to extend such results to more complex contexts 
[@Ovaskainen2010 provides one excellent review], in
particular addressing auto-correlated noise [e.g. @Ripa; @Ripa1999; 
@Petchey; @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
[e.g. @Durrett1994; @Hastings1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
among other issues. We will turn our attention
away from the well-trodden ground of persistence and coexistence to focus
on other phenomena that can emerge in stochastic models: quasi-cycles and
stochastic switching.  Both examples will illustrate patterns that differ
sharply from those of the deterministic skeleton, both for different underlying
reasons.  Interesting in their own right, these examples will also provide
intuition for our third section on noise the informer.

@@

Despite the nearly ubiquitous association of a ball-in-cup model
with ecological discussions of resilience and stability, this
potential well diagram (Figure \ref{noisy_switch}C) is also a source of confusion
on the topic.  As we have seen, the most salient feature of this
diagram is the depth of each well: transition probabilities (under
small noise) depend only on the depth of the well and not the
steepness of the well, the distance between the wells or between
well and tipping point barrier.  However, it is important to remember
that because the potential function is defined by the above integral,
it in fact already combines two more elementary and mechanistically
precise quantities from the growth equation: the slope of the growth
rate (stability) and the distance to the tipping point (size of the
basin of attraction).  Both a steeper negative slope or a longer
distance to integrate over between stable and unstable points will
create a deeper well.
Because these transitions between alternative stable states appear as
such a dramatic deviation from the previous pattern (as illustrated
in Figure \ref{noisy_switch}A), it is  tempting to assume they
result from some sudden external force or change in the
environment. As the numerical examples illustrate,
these transitions can be explained entirely by chance
[@Boettiger2012;  @Drake2013].

@@

The early warning signals approach views noise as 
a sequence of countless miniature perturbation experiments.
Figure \ref{tipping} illustrates early warning indicators in an
experimental manipulation and corresponding simulated model of
@Dai2012 and @Dai2015.  Though early warning signals have been
reported in a growing range of systems [@Scheffer2015], these
laboratory experiments in population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cerevisiae_ on a sucrose medium provide the most convincing 
mechanistic interpretation of the underlying fold bifurcation.
Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation which is accompanied
by the patterns of increasing variance and increasing auto-correlation.
This pattern can be seen empirically in lab populations [@Dai2012; @Dai2015]
and explored numerically in a simulation of the yeast dynamics shown in 
Figure \ref{tipping}. As the mechanistic model of yeast population
dynamics is already quite simple, the simulation shown here (code in Appendix B)
corresponds directly to the model proposed by @Dai2015.

@@

The potential for stochastic phenomena to provide information about underlying
processes goes well beyond available examples in stability and early warning signals.
The examples from fold bifurcations already suggest
that stochastic processes could play a similar role in identifying
other critical transitions. For instance, stochastic resonance may be 
visible in advance of a Hopf bifurcation, where a stable state gives
way to stable oscillations (limit cycle) of growing magnitude
[e.g. @Fussmann2000; @Nelson2013; @Strogatz2001].  Noise can provide
information in unrelated ways as well. in deriving expressions for the variance created by
demographic noise above, we observed that different partitions of
a logistic growth model can lead to the same macroscopic dynamics
(same values for $b$ - $d$), while having different variance (due
to different values for $b$+$d$).  This suggests that the variation
observed in such a population could potentially help resolve
how dynamics are partitioned between individual birth and death
rates (though in practice one would have to first account for any
contribution of environmental noise).

@@

This review has highlighted contrasting views of the role of noise
between specialists and non-specialists and pointed the way to bridging
that divide by seeing noise not as a new source of information.
Most ecologists are introduced to the concept of stochastic models 
through the lens of statistical inference, in which noise is some
relatively abstract nuisance to be stripped away by clever statistical
inference [@Gotelli2004; @Bolker2007].  As we have seen, specialists have long viewed noise in a 
different light, emerging from the same underlying mechanisms that
create both what we think of as the deterministic processes of interest
and the additional fog of noise [@vanKampen2007; @Black2012].  An intuitive approximation of
larger system size both justifies the convenient partition between
a so-called deterministic skeleton and noise components of a model,
while also illustrating how different demographic and environmental
processes will create different noise dynamics [@Coulson2004; @Ovaskainen2010;
@Schreiber2017].  We have also seen the divide in how noise is perceived
goes well beyond merely a more mechanistic origin story. Specialists
have focused heavily on the role of noise as creator of phenomena:
qualitative differences in the behavior of fully stochastic models and
their deterministic skeletons [@Coulson2004].  Tipping our hat to 
the extensive examples in persistence and coexistence, we explored 
examples of quasi-cycles and stochastic switching showing patterns
wholly unlike their deterministic skeletons. Finally, we endeavor to
bring the divergent views of noise in empirical and theoretical communities
together by turning the noise the nuisance on its head: drawing on
examples from the early warning signals literature which have demonstrated
how noisy phenomena can be used to reveal information about slow
changes in the underlying deterministic skeleton.

Divergent views of the role of stochasticity have separated our community
for too long. The eruption of interest in early warning signals for critical
transitions [@Scheffer2009] provides an excellent illustration of
one way to bridge that divide and the value of greater dialogue between
these two communities.  I hope this review challenges specialists to explore
the potential of stochastic phenomena to inform on underlying processes --
such results are more immediately applicable to empirical work that must
infer process from pattern.  Meanwhile, non-specialists could also
benefit from a more nuanced view of stochasticity as a part of the processes
we seek to understand as well as a novel source of information.  All though
mathematical complexity has frequently been a barrier, computational
advances and the widespread
adoption of computational tools such as the R language have made it possible
to explore these phenomena numerically. Unlike static charts
empirical data, simulations can be manipulated to build
an intuition and generate sample data of these processes. To this end, I 
have provided simple and documented code in Appendix B for each of
examples discussed here.  In the past two
decades, we have set to rest those deterministic skeletons that see
noise only as nuisance and embraced the central role noise plays
as the creator of ecological phenomena. Going forward, I hope and
predict that we will see greater discussion and use of stochastic
phenomena to infer underlying processes in empirical patterns in
similar interactions between theoretical and empirical research.
